Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study

### 2025-06-07 inst/paper/main.tex
Edits to introduction

\title{Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks}
\author{Joshua W. Lambert, Adam J. Kucharski}
\date{}

@@

Infectious disease outbreaks pose an everpresent public health threat. The COVID-19 pandemic caused over 7 million deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}. Other pathogens, including those less transmissible, with longer delays between infections, and with less presymptomatic transmission, can also cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}. Thus, timely and accurate response is critical early in an outbreak when the pandemic or epidemic potential is still uncertain \citep{kucharskiControllingMinorOutbreaks2024}. An effective response requires understanding the efficacy and feasibility of intervention strategies across pathogens with varying epidemiological characteristics \citep{fraserFactorsThatMake2004}. \\

The likelihood of an outbreak causing sustained transmission in a population is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical. A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases  per primary case. If the reproduction number is greater than 1, the incidence of infections is likely to increase, and intervention is requireed to cease transmission in a population. If a pathogen is subcritical (i.e. $R < 1$), incidence will decline and eventually stop spreading, but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}. Individual-level heterogeneity in transmission also influences the possibility a pathogen will fail to maintain transmission due to stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}. Higher individual-level variation for a given reproduction number leads to, on average, more outbreaks being unsustained (however a small number of outbreaks may become large due to superspreading). In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}. \\

A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak --- especially in the case of a novel pathogen --- NPIs are often the first line of defence, as pharmaceutical options may not yet be available. These interventions vary in scope and specificity, from targeted measures such as isolation and contact tracing, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020). The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness \citep{longiniContainingPandemicInfluenza2005}. For highly transmissible pathogens with short generation times (e.g. influenza), swift and effective interventions are required to achieve containment. Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}. \\

Isolation of symptomatic infections aims to break the chain of transmission by preventing further infection of susceptible individuals. The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to becoming infectious optimal. This can make it challenging to control outbreaks with symptom-based isolation, as for most pathogens infectiousness preceeds prodromal symptoms. Quarantining contacts exposed to an infectious individuals, independent of whether they later become a case, is a control strategy to reduce transmission before or shortly after symptom onset. Contact tracing is the process of finding the contacts of infected individuals. Both isolation and contact tracing are often used NPIs to try and curtail the growth of an epidemic. Forward contact tracing asks infected individuals who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed. Isolation, quarantine, and contact tracing is a targeted approach that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020}. Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}. However, in epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}. \\

Several pre-COVID-19 studies have concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and contact tracing, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}. However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential. The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, influenza subtypes that have low levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 \citep{Ward2024.12.11.24318702} and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}. Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\

>>>>

\title{Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study}
\author{Joshua W. Lambert, Authors TBC, Adam J. Kucharski}
\date{}

@@

Outbreaks of severe novel infectious diseases pose a major public health threat. The COVID-19 pandemic resulted in over 7 million reported deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}. Recent human cases of avian influenza A/H5N1, alongside transmission, illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption [CITE: Antia et al, Nature, 2003 + key papers on H5N1 - see my PLOS Bio piece]. It is therefore crucial to understand the effectiveness and feasibility of different intervention strategies in the face of such infections.

%Other pathogens, including those less transmissible, with longer delays between infections, and with less presymptomatic transmission, can also cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}.  Timely and accurate response is therefore critical early in an outbreak when the pandemic or epidemic potential is still uncertain \citep{kucharskiControllingMinorOutbreaks2024}. An effective response requires understanding the efficacy and feasibility of intervention strategies across pathogens with varying epidemiological characteristics \citep{fraserFactorsThatMake2004}. \\ AK: Removed to improve flow, as this is mentioned in more detail later in intro.

The likelihood of a novel outbreak causing sustained transmission depends both the characteristics of the pathogen and its host population, as well control measures  implemented in response to early cases. A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases by a typical infectious host. If the reproduction number is greater than 1, we would expect the incidence of infections to increase over time. If $R$ is below 1, we would expect incidence to decline on average, but several cases could still occur before transmission ceases \citep{farringtonDistributionTimeExtinction1999}. Two additional factors can influence the probability a pathogen with $R$ above 1 will successfully establish within a population~\citep{lloyd-smithSuperspreadingEffectIndividual2005}. If there is a large degree of individual-level variation in transmission, the majority of infectious individuals will infect nobody, even though a minority will infect many; as a result, most newly introduced outbreaks will fail to take off. In addition, the likelihood of establishment increases if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}. \\ % AK: Note 'primary case' means first in outbreak

Early in a novel outbreak, pharmaceutical options such as treatments and vaccines may not yet be available, leading to reliance on non-pharmaceutical interventions (NPIs) to reduce infection and disease. These interventions include individual-level targeted measures such as isolation and contact tracing, to population-level policies such as venue closures or restrictions on social mixing (Huag et al., 2020) [AK: probably a better reference out there, e.g. Sharma et al, Nature Comms]. The success of such interventions in curtailing transmission before it becomes sustained depends on several factors, including: the type of intervention, timing of introduction, and the intervention's effectiveness \citep{longiniContainingPandemicInfluenza2005}. For highly transmissible pathogens with a short generation times (e.g. pandemic influenza), incidence of new infections can grow quickly, outpacing a response. Moreover, pathogens such as influenza, which have a large proportion of onward transmission prior to symptom onset and have asymptomatic (subclinical) cases, are less easily controlled via isolation of symptomatic cases and tracing and quarantining of their contacts~\citep{fraserFactorsThatMake2004}. \\

%Isolation of symptomatic infections aims to break the chain of transmission by preventing further infection of susceptible individuals. The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to becoming infectious optimal. This can make it challenging to control outbreaks with symptom-based isolation, as for most pathogens infectiousness preceeds prodromal symptoms [AK: reference? This is a strong claim – best to focus on flu I think]. Quarantining contacts exposed to an infectious individuals, independent of whether they later become a case, is a control strategy to reduce transmission before or shortly after symptom onset. Contact tracing is the process of finding the contacts of infected individuals. Both isolation and contact tracing are often used NPIs to try and curtail the growth of an epidemic. Forward contact tracing asks infected individuals who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed. Isolation, quarantine, and contact tracing is a targeted approach that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020}. Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}. However, in epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}. \\ AK: This is probably too much fundamental background for an applied journal – readers will know what isolation and contact tracing is.

Several pre-COVID-19 studies  concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and tracing their contacts, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}. However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential. The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, influenza subtypes that have low levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 \citep{Ward2024.12.11.24318702} and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}. Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\ [AK: These headline fatality risks are likely to be biased: https://kucharski.substack.com/p/how-fatal-is-h5n1-influenza]
